HW2: Exploring Gapminder
===============================
## Michelle Lee
### Sept 22, 2014


### Loading the dataset

**Load the Gapminder data with read.delim() and read.table() without specifying args other than `file`. Do you get the same resulting object in your workspace?**

```{r setup}
# set global chunk options: images will be 7x5 inches
knitr::opts_chunk$set(fig.width=7, fig.height=5)
options(digits = 4)
```

I loaded the data -but got an error message when I try to use `read.table()`.
```{r, error=TRUE}

gDelim <- read.delim("gapminderDataFiveYear.txt")
gTable <- read.table("gapminderDataFiveYear.txt")
```

**If you're not getting the same object, why not?**

This is the minimum amount of code needed for `read.table()` to work. 

```{r}
gTable <- read.table("gapminderDataFiveYear.txt", fill = TRUE)
```

Documentation for `read.table()` says: "If TRUE then in case the rows have unequal length, blank fields are implicitly added." The warning from R Console indicated that line 146 did not have 6 elements. I checked line 147, and sure enough - Bosnia and Herzegovina overlapped onto the 2nd and 3rd columns.

```{r}
gTable[c(143:147),]
```

**Develop the call to `read.table()` that is equivalent to `read.delim().`**

I tried again by adding more detail - but it wasn't enough to make it equivalent. 
```{r}
gTable <- read.table("gapminderDataFiveYear.txt", header = TRUE, sep = "\t", fill = TRUE)
identical(gTable, gDelim)
```
Why not? On closer examination, I saw that the number of observations aren't the same:

```{r}
str(gTable); str(gDelim)
```

Being somewhat of an R beginner, I looked it up on Stack Overflow on [how to compare two dataframes easily.](http://stackoverflow.com/questions/7494848/standard-way-to-remove-multiple-elements-from-a-dataframe) Only Cote d'Ivoire seems to have a problem, most likely because of the apostrophe in its name.

```{r}
`%notin%` <- function(x,y) !(x %in% y) 
gDelim[(gDelim$country %notin% gTable$country),]
```
I modified the `read.table()` command to include this.

How about now?
```{r}
gTable2 <- read.table("gapminderDataFiveYear.txt", header = TRUE, sep = "\t", quote = "\"", fill = TRUE)
identical(gTable2, gDelim)
```

Yay!
To summarize: I found two major differences between `read.table` and `read.delim`: the quote and fill.



### Smell testing the data

**Is it a data.frame, matrix, vector, list? What's it's mode, class? How many variables? How many rows/observations? What flavour is each variable?**

`str` tells me that it is a data frame, with 1704 observations of 6 variables, and the flavour of each variable - country and continent as factors, year as integer, and popuplation, life expectancy, and GDP per captia as numbers. 
```{r}
gDat=gDelim
str(gDat)
```

**Can you get these facts about extent/size in more than one way? Can you imagine different functions being useful in different contexts?**

Yes - we can use other functions to find specific information. With `dim`, I can find the number of rows and columns...
```{r}
dim(gDat)
```

... names of the variables...
```{r}
names(gDat)
```

... the flavour of individual variable...
```{r}
class(gDat$continent)
```

... and a useful summary of all the data.
```{r}
summary(gDat)
```

## Exploring individual variables

**Pick one categorical variable and one quantitative variable to explore. Characterize what's possible (max, min, etc). What's typical? What's the spread? What's the distribution? Feel free to use summary stats, tables, figures.**

I first looked at GDP per capita. Here's a summary of the maximum, minimum, and quantiles:

```{r}
summary(gDat$gdpPercap)
```

However, it's a lot easier (and more fun) to explore using ggplot2, and to play around with aesthetics, such as x, y, alpha, colour, fill, shape, and size. 

```{r}
library(ggplot2)
p <- ggplot(gDat, aes(y = gdpPercap, x = year))
p <- p + scale_y_log10()
p + geom_point(aes(color=continent)) 
```


The huge variation in GDP made it difficult to see any real trends over time. The mix of colours weren't very helpful - so I separated the data by continent. 

```{r}
p + geom_point(alpha = (1/3), size = 3, aes(color=continent)) + facet_wrap(~ continent)
```


I found this to be a better separation of continental trends. Asia seemed to have the most chaotic pattern, so that's what I investigated next.



```{r}
aDat <- subset(gDat, subset=(continent=="Asia"))
asia <- ggplot(aDat, aes(y = gdpPercap, x = year)) + scale_y_log10() 
asia + geom_point(alpha = (1/3), size = 3, aes(color=country)) + facet_wrap(~ country, ncol=6, nrow=6) + theme(legend.position="none")
```


It requires a full screen to fully appreciate the graph, but I love how you can see all trends and countries at once without using a map - otherwise, small countries like Qatar can barely be seen! Andrew Gelman has a post further elaborating on [why lineplots are (sometimes) better than maps](http://andrewgelman.com/2014/04/10/small-multiples-lineplots-maps-ok-always-yes-case/). 

The one thing I tried (and ultimately failed) was to reorder the plots by highest GDP per capita, rather than alphabetically. However, that involved reordering factors, as country is a factor, and I couldn't get it done, despite [Jenny's tutorial](http://www.stat.ubc.ca/~jenny/STAT545A/block08_bossYourFactors.html#how-to-change-factor-levels-reordering) - hopefully I'll realize what I'm doing wrong. 



I went on to look at the life expectancy data. Here's a plot on life expectancy for all countries.

```{r}
le <- ggplot(gDat, aes(y = lifeExp, x = year)) + geom_point(aes(color=continent)); le
```

Here is a closer look at life expectancies within Africa. Life expectancy ranges from 23.60 to 76.44! 
```{r}
africa <- subset(gDat, subset=(continent=="Africa"))
summary(africa)
```

I picked 5 African countries: Botswana, Libya, Mauritius, and Sierra Leone, based on high and low life expectancies. 
```{r}
head(africa[ order(-africa$year, -africa$lifeExp),])
tail(africa[ order(-africa$year, -africa$lifeExp),])
```

I plotted life expectancy over time for these five countries, and added in China for added perspective.

```{r}
countries<-subset(gDat, country=="Sierra Leone" | country=="Libya" | country=="Mauritius" | country=="China" | country=="Botswana")
af <- ggplot(countries, aes(x=year, y=lifeExp, colour=country)) + geom_line(); af
```

This was surprising (to me, at least) that China's life expectancy is so similar to that of Libya and Mauritius. Another noticeable trend is the huge dip in life expectancy between 1990 and 2000 for Botswana, [widely attributed](http://www.aho.afro.who.int/profiles_information/index.php/AFRO:Life_expectancy) to the catastrophic impacts of HIV/AIDS.


Lastly, I made a map of Africa to explore life expectancies on a spatial level. North African countries are clearly better off than their sub-Saharan counterparts. There may be a trend of landlocked countries having lower life expectancies. I might come back to it later - there are plenty of [resources on analysis of spatial data](http://cran.r-project.org/web/views/Spatial.html) in R.

```{r, echo=FALSE}
library(rworldmap)
africa <- subset(gDat, subset=(continent=="Africa") & (year > 2005))
gtdMap <- joinCountryData2Map(africa, nameJoinColumn="country",joinCode="NAME")
```
```{r}
mapCountryData(gtdMap, nameColumnToPlot="lifeExp", mapRegion = "africa",catMethod="fixedWidth", numCats=100, mapTitle="Life Expectancy in Africa (2007)")
```



### Report your process
* Playing around with ggplot2 was a great learning experience - I learned several formats in which certaom data could be presented easily and effectively, especially the line plots. 
* As I mentioned earlier - reordering factors was the most frustrating task, and one I didn't manage to solve. 
* Helpful tutorials I used include: Hadley Wickham's [documentation for ggplot](http://docs.ggplot2.org/0.9.3.1/geom_bar.html) and how to create a [chloropleth on rworldmap](http://stackoverflow.com/questions/22625119/choropleth-world-map).

```{r compile, eval=FALSE}
library(knitr)
knit('HW2-gapminder.Rmd')
```

