---
title: 'HW 5: Factors and Files'
author: "Michelle Lee"
date: "October 16, 2014"
output:
  html_document:
    keep_markdown: yes
    keep_md: yes
    toc: yes
    toc_depth: 2
---

## Prior tasks
```{r}
library(ggplot2)
library(ggthemes)
library(knitr)
library(reshape2)
library(plyr)
suppressPackageStartupMessages(library(dplyr))
gDat <- read.delim("gapminderDataFiveYear.txt")
gDat <- tbl_df(gDat)
```

## Drop Oceania

I used `droplevels` to remove unused factor levels. 
```{r}
hDat <- gDat %>%
  filter(!(continent %in% "Oceania")) %>%
  droplevels
```

We can compare the data before and after dropping Oceania, such as number of rows, number and identity of continents and countries.
```{r, results='asis'}
table<-matrix(ncol=2, nrow=3)
table<- rbind(cbind(nrow(gDat), nrow(hDat)), cbind(length(unique(gDat$continent)), length(unique(hDat$continent))), cbind(as.character(unique(gDat$continent)), as.character(unique(hDat$continent))))
table[7,2]<-""
rownames(table) = c("Number of rows", "Number of continents", "Continents", "", "" ,"" ,"")
colnames(table) = c("Before", "After")
knitr::kable(table)
```

Above is a table with the descriptive statistics summarized.



## Model life expectancy as a function of year 

I used the code Jenny gave us.

```{r}
j_coefs <- ddply(gDat, ~ country + continent, function(dat, offset = 1952) {
  the_fit <- lm(lifeExp ~ I(year - offset), dat)
  setNames(coef(the_fit), c("intercept", "slope"))
})
```

Now we can look at the properties of `j_coefs`. 

```{r}
str(j_coefs)
```

This tells us about the structure of `j_coefs` and what sort of data is provided, but it may be more interesting to see which countries have the highest and lowest levels. 

```{r, results='asis'}
knitr::kable(j_coefs %>%
  arrange(desc(slope)) %>%
  head)
```

Now we can see the country with the highest slope is Oman. If we did it by intercept:

```{r, results='asis'}
knitr::kable(j_coefs %>%
  arrange(desc(intercept)) %>%
  head)
```

Norway is the country with the highest intercept. Again, not particulary surprising.


## Order of data vs. order of factor levels

Jenny's code is given below:
```{r}
post_arrange <- j_coefs %>% arrange(slope)
post_reorder <- j_coefs %>%
  mutate(country = reorder(country, slope))
post_both <- j_coefs %>%
  mutate(country = reorder(country, slope)) %>%
  arrange(country)
```

Let's look at all three of these at once.

```{r}
head(post_arrange); head(post_reorder); head(post_both)
```

`post_arrange` and `post_both` look the same. We can look at these one by one.

### post_arrange

```{r}
head(post_arrange); tail(post_arrange)
```

We can clearly see that `post_arrange` sorted the data according to increasing order of slope. 

### post_reorder

```{r}
head(post_reorder); tail(post_reorder)
```

Here, we can see that `country` is organized alphabetically first, and then slope after. Because `mutate` is used, we know that the `country` factor has been reordered. 

```{r, results='asis'}
table2<- head(cbind((levels(j_coefs$country)),(levels(post_reorder$country)), as.data.frame(post_reorder$slope)), n=10)
colnames(table2) = c("Levels before", "Levels after", "Slope")
knitr::kable(table2)
```

Now we can see that alphabetical order isn't how the levels are ordered now - it is by slope. We know that it's not just simple sorting, but an actual change in levels, through ggplot. I'll use the chart Jenny showed us in class:

```{r}
ggplot(post_arrange, aes(x = slope, y = country)) + geom_point(size = 1.2, colour="purple") + 
      ggtitle ("Slope order and factors - no change") + guides(fill = F) + 
  theme(plot.title = element_text(lineheight=.8, face="bold", size = 15)) + 
  xlab("Country") + ylab("Slope")
```

This is `post_arrange`, where the data was sorted but **not** with reordered factor levels. Below is `post_reorder`, which is clearly reordered:

```{r}
ggplot(post_reorder, aes(x = slope, y = country)) + geom_point(size = 1.2, colour="dark blue") + 
      ggtitle ("Slope order and factors - no change") + guides(fill = F) + 
  theme(plot.title = element_text(lineheight=.8, face="bold", size = 15)) + 
  xlab("Country") + ylab("Slope")
```

### post_booth

`post_booth` combines both previous objects; it mutates the country level first, and then arranges it by country, which is why it looked identical to `post_arrange` (although by no means the same!) We can convince ourselves of this by checking:

```{r}
identical(post_both, post_arrange)
```

Great! Now we can check the levels:
```{r, results='asis'}
table3<- head(cbind((levels(post_arrange$country)),(levels(post_both$country))), n=10)
colnames(table3) = c("Levels for post_arrange", "Levels for post_both")
knitr::kable(table3)
```

We can see that the two are again very different, because of the reordered factors. What about `post_reorder`?

```{r, results='asis'}
table3<- head(cbind((levels(post_reorder$country)),(levels(post_both$country))), n=10)
colnames(table3) = c("Levels for post_reorder", "Levels for post_both")
knitr::kable(table3)
```

The levels are the same...
```{r}
identical(post_reorder, post_both)
```

... but they are not identical. This can be easily checked again:

```{r}
head(post_reorder); head(post_both)
```

Again, it is clear that `post_both` rearranges the levels like `post_reorder`, but arranges the country order to look like `post_arrange`. 

### In the definition of post_both, if I swapped out arrange(country) for arrange(slope), would I get the same result?

```{r}
pb2<- post_both %>%
  arrange(slope)
identical(post_both, pb2)
```

Yes, they are identical, because `arrange` is simply changing how the data is presented, but nothing else at a deeper level. 

### What if the reorder() command got changed?

Let's try changing it to `reorder(country, intercept)`. 

```{r, results='asis'}
pb2<- j_coefs %>%
  mutate(country = reorder(country, intercept)) %>%
  arrange(country) 
table4<- head(cbind((levels(post_both$country)),(levels(pb2$country)), as.data.frame(pb2$intercept)), n=10)
colnames(table4) = c("Levels for post_both", "Levels for post_both changed", "PB changed intercept")
knitr::kable(table4)
```

It's clear that the changed post_booth is by intercept. 


### Do you have any preference for one arrange statement over the other?

It really depends on which factor you want to concentrate on. `ggplot` doesn't plot well without factors being reordered, so it would make sense to reorder the factor of interest. 


## Revalue a factor 

```{r}
countries<- c("Korea, Rep.", "United States", "Japan", "Canada", "India", "Iceland")
countries2 <- gDat %>%
  filter(country %in% countries) %>%
  droplevels
new_countries<-c("Korea, Rep." = "Where I was born", "United States" = "Where I went next", "Japan" = "Where I went after that", "Canada" = "Where I went to high school", "India" = "Where I went backpacking and got lost", "Iceland" = "Where I hope to explore fjords")
places <- revalue(countries2$country, new_countries)
```

To make sure that our factors have been revalued:

```{r, results='asis'}
mDat <- gDat %>%
  filter(country %in% countries) %>%
  mutate(new_factor = revalue(countries2$country, new_countries)) %>%
  select(country, new_factor)
knitr::kable(unique(mDat))
```

The levels have been changed! 


## Reorder a factor

`ggplot` doesn't allow you to change the order in which countries are plotted, unless the factor is changed. Here, I changed the factor level such that the countries are organized by life expectancy in 1952, in ascending order.

```{r, fig.width=8, fig.height=8}
asia <- gDat %>%
  mutate(country = reorder(country, lifeExp))  %>% # reorder by life expectancy! 
  filter(continent=="Asia") %>%
  select(country, year, lifeExp) %>%
  droplevels

ggplot(asia, aes(x = year, y = lifeExp, colour = lifeExp < quantile(asia$lifeExp, 0.25))) + # by 1st quartile life expectancy
  geom_point(aes(group=year)) + facet_wrap(~ country) + 
  ggtitle("Life expectancy and continential 1st quantile over time") + 
  theme(legend.position="none", plot.title = element_text(size = 15, face="bold")) + 
  xlab("Year") + ylab("Life Expectancy") + scale_color_economist()
```

The data has now been reordered, so that the plot starts with the country with the lowest life expectancy in 1952 (Afghanistan) to the highest (Japan). The point when a country turns from dark to light blue indicates when they were no longer within the 1st quantile of life expectancy.


## Write and read data to file


### write.table, read.table


First, let's test write and read.table.

```{r}
# export and import using test data
countries<- c("Afghanistan", "Canada", "Iceland", "Japan", "Sweden")
testData <- gDat %>% 
  filter(country %in% countries)  %>%
  as.data.frame %>%
  droplevels
write.table(testData, "testData.csv", sep=" ", row.names=F, quote=F)
check.table<- read.csv("testData.csv", sep=" ", header=T)
identical(testData, check.table)
```

The two aren't the same. If we take a closer look:

```{r}
str(testData); str(check.table)
```

The difference is that population is an integer for `check.table`, instead of number for `testData`. If we try fixing that:

```{r}
check.table$pop<-as.numeric(check.table$pop)
identical(testData, check.table)
```

Sure enough, they are identical now!


### RDS
 
Let's try saving as an RDS object.

```{r}
saveRDS(testData, "testData.rds")
# import and try testing it
checkRDS<- readRDS("testData.rds")
identical(testData, checkRDS)
```

Exporting and re-importing an RDS file is significantly easier than it is for a csv - with write and read.table, you have to worry about whether strings have commas, the class of numbers, and so on. 


### dput

```{r}
dput(testData,"testData.txt")
# import it back 
checkDGet <- dget("testData.txt")
identical(testData, checkDGet)
```

Again, it is identical.

### Which is better?

After trying all three methods, I think RDS is the most convenient. It keeps the level order and there is less to worry about the data behaving badly when it is exported and imported it back. 

### Naming conventions?

If it is a file I will be accessing frequently, I often name it by date, because it is easiest for me to pick out from a list of files. However, if it is a file I will be returning to much later, then I would rename it to something more descriptive. Now that I've learned about RDS objects, I'll be definitely using it more, rather than exporting to csv's!


## Report your process

* Reordering the factors was easy once I learned the methods. It's wonderful how convenient `dplyr` is, and how easy it makes organizing and munging data.
* Difficult: exporting and importing csv's, especially if the country names have commas and other punctuation that bring all sorts of grief. RDS files are so much better to work with.